Methods for Analyzing Models of\\ Rod-Shaped Bacteria

### 2025-10-24 paper/main.tex
rename section to make purpose more clear
reformulate cost function sections

% --------------------------------------------------------------------------------------------------
\subsection{Data Extraction}

@@

\paragraph{No Cell Division}
In the scenario where no cells divide during the observed interval, we can use our extraction
algorithm to obtain the position of the cells and then compare them directly to the positions
calculated from the computational results. For all cells $j=1..M$, we have a collection of $i=0..N$ vertices $\vec{x}_{i,j}$. \textbf{[Warum jetzt zweifach indiziert?]} Thus, the total amount of information is an array of shape $(M,N,2)$ Assuming that we have $N$ cells, each of which is represented by $M$ vertices, the total amount of information is an array of shape $(N,M,2)$, containing all positions of all agents.
It is thus only a matter of comparing two of these arrays with each other and many methods and
approaches which already exist~\cite{Wang2020} can be leveraged. For now, we simply calculate the distance of each vertex pair, which is equivalent to the
well-known loss function:

\begin{equation}
    L_{\text{LS}}(\vec{x}, \vec{y})
        = \sqrt{\sum\limits_{i,j}||\vec{x}_{i,j}-\vec{y}_{i,j}||^2}
        = \sqrt{\sum\limits_{i,j,k}|x_{i,j,k}-y_{i,j,k}|^2}
\end{equation}

of least squares (LS).
One important detail is that it must be guaranteed that the ordering of vertices is identical in the
numerical and experimentally obtained results.

\paragraph{Comparison across Generations}
In order to compare a variable amount of cells, we can not utilize the positions of these cells
for comparison directly since one particular state of our simulation might contain cells which are
not present in the experimental results (or vice-versa).
\textbf{[Es bleibt unklar, was du hier verhandeln möchtest.]}

We thus need additional methods which can be used for comparisons across division events.
To achieve this, we use the numerical results of the \ac{abm} simulation and our visualization
algorithms to generate cell-masks (see section~\ref{subsection:visualization}).
We then compare the generated masks to the experimentally obtained ones pixel by pixel.
Whenever two pixel colors are not matching, a loss value is assigned.
The sum of all those penalties is then the total loss.
This algorithm is an implementation of a signed area measure.
It is very important to take into account the parental relationship between the cells, checking if
either one of the cells is a daughter of the other cell.
Figure~\ref{fig:mask-difference-metric} (A,B) shows two successive states of a numerical simulation
with varying amounts of cells.
We can see that some division events have occurred but other cells have only grown without
completing a division step.
The resulting penalties without accounting for the parental relationships are displayed in
Figure~\ref{fig:mask-difference-metric} (C).
Black indicates that the pixels are matching and no loss is assigned while a white pixel
indicates that the original pixels are not matching, thus assigning a loss value.
Similarly, Figure~\ref{fig:mask-difference-metric} (D) takes the parental relationship into account
and highlights these in gray.
It is possible to assign a custom value for the parental loss value.
If we set this loss value to $0$, this algorithm ensures the contiuity of the loss function in time.

But it is also biologically motivated.
If a cell has just divided, the resulting two daughter cells will cover precisely the space that the
mother had taken up.
Often times, the exact time at which the fission event is over can not be determined accurately,
thus leading not only to uncertainties but also some fundamental problems.
Another approach to construct a comparison algorithm across generations would be to use the
previously explained position comparison algorithm and use it on intervals in which no division
event occurrs.
But this attempt would quickly lead to multiple problems.
How can we deal with the uncertainties of the timing of the division event?
And the more cells we want to investigate, the more division events will occurr even in a fixed
amount of time thus greatly decreasing the effective interval at which this alternative algorithm
could be used.
Furthermore, what if there are only two images and there is a division event which is occurring
between them?

It further needs to be noted that the constructed algorithm can also be adapted by users in order to
provide more complex loss functions (such as considering time-evolvement, etc.).
Our method can make use of existing machine leanring algorithms since it is only based on the
principle of comparing two images with each other.
There already exists a variety of methods on how to assign loss values in these
circumstances~\cite{Dice1945}.

@@

\begin{figure}
    \centering
    \begin{minipage}{0.5\textwidth}
    \begin{tikzonimage}[width=0.48\textwidth]
        {docs/source/_static/fitting-methods/progressions-1.png}
        \node at (0.025, 0.975)[anchor=north west, rectangle, draw, white, minimum width=15pt, minimum height=15pt]{\textbf{A}};
    \end{tikzonimage}%
    \hspace{0.01\textwidth}%
    \begin{tikzonimage}[width=0.48\textwidth]
        {docs/source/_static/fitting-methods/progressions-2.png}
        \node at (0.025, 0.975)[anchor=north west, rectangle, draw, white, minimum width=15pt, minimum height=15pt]{\textbf{B}};
    \end{tikzonimage}
    \linebreak
    \vspace{0.01\textwidth}
    \begin{tikzonimage}[width=0.48\textwidth]
        {docs/source/_static/fitting-methods/progressions-3.png}
        \node at (0.025, 0.975)[anchor=north west, rectangle, draw, white, minimum width=15pt, minimum height=15pt]{\textbf{C}};
    \end{tikzonimage}%
    \hspace{0.01\textwidth}%
    \begin{tikzonimage}[width=0.48\textwidth]
        {docs/source/_static/fitting-methods/progressions-4.png}
        \node at (0.025, 0.975)[anchor=north west, rectangle, draw, white, minimum width=15pt, minimum height=15pt]{\textbf{D}};
    \end{tikzonimage}
    \end{minipage}%
    \begin{minipage}{0.49\textwidth}
        \begin{tikzonimage}[width=\textwidth]
            {docs/source/_static/fitting-methods/penalty-time-flow.png}%
            \node at (0, 0.975)[anchor=north west, draw, rectangle, minimum width=15pt, minimum height=15pt]{\textbf{E}};
        \end{tikzonimage}
    \end{minipage}
    \caption{
        The figure shows two masks (A,B) which will be compared against each other.
        Image (B) contains more cells which are a result of cellular division events that have not
        yet occurred in image (A).
        (C) "Naive" comparison of cell masks, which identifies pixels of non-matching colors and
        assigns a loss by counting these pixels.
        This simple approach does not take into consideration the parental relationship.
        This approach is displayed in (D) where pixels of cells that are directly related have been
        colored in gray.\\
        (E) Comparison of loss functions, exponential regression (ER) and number of cells.
        % Total loss between successive iteration steps $t_n,t_{n+1}$.
        The overall shape of the loss is expected to have the slope of an exponential function.
        We can clearly see that division events introduce jumps within this relationship.
        In the case where we account for the parental relationship, these jumps are mitigated.
        The loss value was set to $0$ for pixel-comparisons which belong to cells that are in a
        parental relationship.
    }
    \label{fig:mask-difference-metric}
\end{figure}

\paragraph{Testing the Comparison Algorithm with Division Events}
When utilizing a computational model (see section~\ref{sec:computational-model}), we can generate
data and use the generated masks to provide a benchmark of our algorithm.
To do this, we set out to calculate the loss between two successive iteration steps
$t_n$ with $t_{n+1}$.
We expect to obtain an exponential growth in the loss.
This can be explained by the fact that the growth of the total area is exponential in time.
Furthermore, the temporal area difference can be viewed as a derivative of the total area in time,
which is again an exponential function.
Figure~\ref{fig:mask-difference-metric} (E) shows the total loss for this approach.
We can see that by accounting for the parental relationship, we get the overall desired effect.
This method allows us to compare results between states that have a different number of cells.
However, since this algorithm requires us to calculate a cell-mask for each comparison and then also
process this mask as well as the comparator, it is much more computationally demanding.
After all, for each comparison step, we need to run a full 3D visualization pipeline.
In cases where the simpler approach of comparing positions of the agents is possible, this method
should be preferred.
\textbf{[Wir müssen darüber reden, mir ist hier vieles unklar. Es könnte nützlich sein, vor der Application Sektion, eine Zusammenfassung zu setzen.]}

@@

By doing so, we can utilize the simpler and more performant position comparison scheme for our loss function.
\textbf{[Hier weiss man nicht worüber du sprichst.]}

@@

\begin{figure}
    \centering
    \begin{tikzonimage}[width=0.25\textwidth]
        {figures/crm_fit/profiles/damping.pdf}%
        \node at (0, 1.1)[anchor=north west, rectangle, draw, black, minimum width=15pt, minimum height=15pt]{\textbf{A}};
    \end{tikzonimage}%
    \begin{tikzonimage}[width=0.25\textwidth]
        {figures/crm_fit/profiles/exponent-m.pdf}%
        \node at (0, 1.1)[anchor=north west, rectangle, draw, black, minimum width=15pt, minimum height=15pt]{\textbf{B}};
    \end{tikzonimage}%
    \begin{tikzonimage}[width=0.25\textwidth]
        {figures/crm_fit/profiles/exponent-n.pdf}%
        \node at (0, 1.1)[anchor=north west, rectangle, draw, black, minimum width=15pt, minimum height=15pt]{\textbf{C}};
    \end{tikzonimage}%
    \begin{tikzonimage}[width=0.25\textwidth]
        {figures/crm_fit/profiles/strength.pdf}%
        \node at (0, 1.1)[anchor=north west, rectangle, draw, black, minimum width=15pt, minimum height=15pt]{\textbf{D}};
    \end{tikzonimage}\\
    \begin{tikzonimage}[width=0.25\textwidth]
        {figures/crm_fit/profiles/radius.pdf}%
        \node at (0, 1.1)[anchor=north west, rectangle, draw, black, minimum width=15pt, minimum height=15pt]{\textbf{E}};
    \end{tikzonimage}%
    \begin{tikzonimage}[width=0.25\textwidth]
        {figures/crm_fit/profiles/growth-rate-0.pdf}%
        \node at (0, 1.1)[anchor=north west, rectangle, draw, black, minimum width=15pt, minimum height=15pt]{\textbf{F}};
    \end{tikzonimage}%
    \begin{tikzonimage}[width=0.25\textwidth]
        {figures/crm_fit/profiles/growth-rate-1.pdf}%
        \node at (0, 1.1)[anchor=north west, rectangle, draw, black, minimum width=15pt, minimum height=15pt]{\textbf{G}};
    \end{tikzonimage}%
    \begin{tikzonimage}[width=0.25\textwidth]
        {figures/crm_fit/profiles/growth-rate-2.pdf}%
        \node at (0, 1.1)[anchor=north west, rectangle, draw, black, minimum width=15pt, minimum height=15pt]{\textbf{H}};
    \end{tikzonimage}%
    \caption{
        We calculated the likelihood profiles for every optimized parameter of our model.
        All plots show how the value of the loss function varies when fixing a single parameter but
        still optimizing for the remaining ones.
        Every plot indicates that a minimum was achieved, thus every parameter could be identified.
        We show only one of the 6 individually-chosen growth rate as an example.
        The remaining plots show similar behaviour although with varying values and can be found in
        the supplementary material in section~\ref{sec:supplement-parameter-profiles}.
    }
    \label{fig:parameter-estimates-single-step}
\end{figure}

>>>>

% --------------------------------------------------------------------------------------------------
\subsection{Position Extraction}

@@

\paragraph{Direct Comparison}
In the scenario where no cells divide during the observed interval, we can use our extraction
algorithm to obtain the position of the cells and then compare them directly to the positions
calculated from the computational results.
Each cell contains a collection of vertices.
Given $M$ cells with $N$ vertices each and dimension $d$, the total information can be stored in an
array $\nu_{i,j,k}$ of shape $(M,N,2)$.
Comparing the results of a simulation $\nu$ with experimental values $\tilde{\nu}$ is thus only a
matter of comparing two of these arrays within a suitable cost function~\cite{Wang2020}.
In our approach we utilize a least squares approach given by

\begin{equation}
    L_{\text{LS}}(\vec{x}, \vec{y})
        = \sqrt{\sum\limits_{i,j,k}|\nu_{i,j,k} - \tilde{\nu}_{i,j,k}|^2}
\end{equation}

which minimizes the distance between the corresponding vertices.
One important detail is that the ordering of vertices must be identical in the numerical and
experimental values.

\paragraph{Comparison across Generations}
The preceding algorithm is performant and effective but can not be applied in the event of a
division process.
Due to possibly differing division times, a cell which is present in the simulation might not yet
have a corresponding cell in the experimental data at this time point yet (or vice-versa).
In order to still be able to compare these datapoints, we require additional methods which can be
used for comparisons across division events.
\textbf{[Es bleibt unklar, was du hier verhandeln möchtest.]}
\textit{Jetzt klarer?}

To this end we use the numerical results of the \ac{abm} simulation and our visualization
algorithms to generate cell-masks (see section~\ref{subsection:visualization}).
Each pixel has a unique color value which is directly related to one particular cell or background
(black).
We compare the generated masks to the experimentally obtained ones pixel by pixel.
Whenever two colors are not matching, a penalty of $1$ is assigned.
The total cost is the sum of all penalties.
% This algorithm is an implementation of a signed area measure.
It is important to take into account the parental relationship between the cells, by checking if
either one of the colors corresponding to the compared colors is a daughter of the other cell.
Figure~\ref{fig:mask-difference-metric} (A,B) shows two successive states of a numerical simulation
with varying amounts of cells.
We can see that some cells have divided while others have only grown without completing a division
step.
The resulting penalties without accounting for the parental relationships are displayed in
Figure~\ref{fig:mask-difference-metric} (C).
Black indicates that the pixels are matching and no penalty is assigned while a white pixel
indicates that the original pixels are not matching, thus adding to the overall cost.
In contrast, Figure~\ref{fig:mask-difference-metric} (D) takes the parental relationship into
account and highlights regions where compared pixels belong to mother-daughter cells in gray.
We assign a penalty $0\leq p \leq 1$ which is a hyperparameter for this optimization algorithm.
% It is possible to assign a custom value for the parental cost value.
Settings its value to $0$ results in a contiuous cost function in time.

Future extensions should introduce a time-dependent parent penalty $p=p(t)$ which grows with the age
of the daughter cells.
Furthermore, the developed algorithm relies on the comparison of two images and can as such make use
of existing machine-learning workflows that construct cost functions for such
scenarios~\cite{Dice1945}.\\
\textbf{@Christian are these two sentences Discussion?}

Although this optimization scheme was driven by technical challenges, it is also biologically
motivated.
In the infenitisemal time point after the division event the resulting two daughter cells will cover
precisely the space that the mother had taken up.
Often times, the exact time at which the fission event has commenced can not be determined
accurately, thus leading to uncertainties with respect to the question when the life of the daughter
cells begins and the one of the mother ends.

In a different approach, one might be tempted to utilize the previously constructed direct
comparison algorithm for time intervals that do not contain any division events.
But this attempt would quickly lead to multiple problems.
It will not be able to deal with the uncertainties of the timing of the division event.
As the number of investigated cells grows, more division events will occurr even in a fixed
amount of time thus greatly decreasing or completely vanishing the effective interval at which no
division event is observed.
The latter situation is equivalent to two microscopic images as input data between which a division
event has occured.

\textbf{@Christian is this Supplement?}

@@

\begin{figure}
    \centering
    \begin{minipage}{0.5\textwidth}
    \begin{tikzonimage}[width=0.48\textwidth]
        {docs/source/_static/fitting-methods/progressions-1.png}
        \node at (0.025, 0.975)[anchor=north west, rectangle, draw, white, minimum width=15pt, minimum height=15pt]{\textbf{A}};
    \end{tikzonimage}%
    \hspace{0.01\textwidth}%
    \begin{tikzonimage}[width=0.48\textwidth]
        {docs/source/_static/fitting-methods/progressions-2.png}
        \node at (0.025, 0.975)[anchor=north west, rectangle, draw, white, minimum width=15pt, minimum height=15pt]{\textbf{B}};
    \end{tikzonimage}
    \linebreak
    \vspace{0.01\textwidth}
    \begin{tikzonimage}[width=0.48\textwidth]
        {docs/source/_static/fitting-methods/progressions-3.png}
        \node at (0.025, 0.975)[anchor=north west, rectangle, draw, white, minimum width=15pt, minimum height=15pt]{\textbf{C}};
    \end{tikzonimage}%
    \hspace{0.01\textwidth}%
    \begin{tikzonimage}[width=0.48\textwidth]
        {docs/source/_static/fitting-methods/progressions-4.png}
        \node at (0.025, 0.975)[anchor=north west, rectangle, draw, white, minimum width=15pt, minimum height=15pt]{\textbf{D}};
    \end{tikzonimage}
    \end{minipage}%
    \begin{minipage}{0.49\textwidth}
        \begin{tikzonimage}[width=\textwidth]
            {docs/source/_static/fitting-methods/penalty-time-flow.png}%
            \node at (0, 0.975)[anchor=north west, draw, rectangle, minimum width=15pt, minimum height=15pt]{\textbf{E}};
        \end{tikzonimage}
    \end{minipage}
    \caption{
        The figure shows two masks (A,B) which will be compared against each other.
        Image (B) contains more cells which are a result of cellular division events that have not
        yet occurred in image (A).
        (C) "Naive" comparison of cell masks, which identifies pixels of non-matching colors and
        assigns a cost by counting these pixels.
        This simple approach does not take into consideration the parental relationship.
        This approach is displayed in (D) where pixels of cells that are directly related have been
        colored in gray.\\
        (E) Comparison of cost functions, exponential regression (ER) and number of cells.
        % Total cost between successive iteration steps $t_n,t_{n+1}$.
        The overall shape of the cost is expected to have the slope of an exponential function.
        We can clearly see that division events introduce jumps within this relationship.
        In the case where we account for the parental relationship, these jumps are mitigated.
        The cost value was set to $0$ for pixel-comparisons which belong to cells that are in a
        parental relationship.
    }
    \label{fig:mask-difference-metric}
\end{figure}

\paragraph{Testing the Comparison Algorithm with Division Events}
When utilizing a computational model (see section~\ref{sec:computational-model}), we can generate
data and use the generated masks to provide a benchmark of our algorithm.
To do this, we set out to calculate the cost between two successive iteration steps
$t_n$ with $t_{n+1}$.
We expect to obtain an exponential growth in the cost.
This can be explained by the fact that the growth of the total area is exponential in time.
Furthermore, the temporal area difference can be viewed as a derivative of the total area in time,
which is again an exponential function.
Figure~\ref{fig:mask-difference-metric} (E) shows the total cost for this approach.
We can see that by accounting for the parental relationship, we get the overall desired effect.
This method allows us to compare results between states that have a different number of cells.
However, since this algorithm requires us to calculate a cell-mask for each comparison and then also
process this mask as well as the comparator, it is much more computationally demanding.
After all, for each comparison step, we need to run a full 3D visualization pipeline.
In cases where the simpler approach of comparing positions of the agents is possible, this method
should be preferred.
\textbf{[Wir müssen darüber reden, mir ist hier vieles unklar. Es könnte nützlich sein, vor der Application Sektion, eine Zusammenfassung zu setzen.]}

@@

By doing so, we can utilize the simpler and more performant direct comparison scheme for our cost function.
\textbf{[Hier weiss man nicht worüber du sprichst.]}

@@

\begin{figure}
    \centering
    \begin{tikzonimage}[width=0.25\textwidth]
        {figures/crm_fit/profiles/damping.pdf}%
        \node at (0, 1.1)[anchor=north west, rectangle, draw, black, minimum width=15pt, minimum height=15pt]{\textbf{A}};
    \end{tikzonimage}%
    \begin{tikzonimage}[width=0.25\textwidth]
        {figures/crm_fit/profiles/exponent-m.pdf}%
        \node at (0, 1.1)[anchor=north west, rectangle, draw, black, minimum width=15pt, minimum height=15pt]{\textbf{B}};
    \end{tikzonimage}%
    \begin{tikzonimage}[width=0.25\textwidth]
        {figures/crm_fit/profiles/exponent-n.pdf}%
        \node at (0, 1.1)[anchor=north west, rectangle, draw, black, minimum width=15pt, minimum height=15pt]{\textbf{C}};
    \end{tikzonimage}%
    \begin{tikzonimage}[width=0.25\textwidth]
        {figures/crm_fit/profiles/strength.pdf}%
        \node at (0, 1.1)[anchor=north west, rectangle, draw, black, minimum width=15pt, minimum height=15pt]{\textbf{D}};
    \end{tikzonimage}\\
    \begin{tikzonimage}[width=0.25\textwidth]
        {figures/crm_fit/profiles/radius.pdf}%
        \node at (0, 1.1)[anchor=north west, rectangle, draw, black, minimum width=15pt, minimum height=15pt]{\textbf{E}};
    \end{tikzonimage}%
    \begin{tikzonimage}[width=0.25\textwidth]
        {figures/crm_fit/profiles/growth-rate-0.pdf}%
        \node at (0, 1.1)[anchor=north west, rectangle, draw, black, minimum width=15pt, minimum height=15pt]{\textbf{F}};
    \end{tikzonimage}%
    \begin{tikzonimage}[width=0.25\textwidth]
        {figures/crm_fit/profiles/growth-rate-1.pdf}%
        \node at (0, 1.1)[anchor=north west, rectangle, draw, black, minimum width=15pt, minimum height=15pt]{\textbf{G}};
    \end{tikzonimage}%
    \begin{tikzonimage}[width=0.25\textwidth]
        {figures/crm_fit/profiles/growth-rate-2.pdf}%
        \node at (0, 1.1)[anchor=north west, rectangle, draw, black, minimum width=15pt, minimum height=15pt]{\textbf{H}};
    \end{tikzonimage}%
    \caption{
        We calculated the likelihood profiles for every optimized parameter of our model.
        All plots show how the value of the cost function varies when fixing a single parameter but
        still optimizing for the remaining ones.
        Every plot indicates that a minimum was achieved, thus every parameter could be identified.
        We show only one of the 6 individually-chosen growth rate as an example.
        The remaining plots show similar behaviour although with varying values and can be found in
        the supplementary material in section~\ref{sec:supplement-parameter-profiles}.
    }
    \label{fig:parameter-estimates-single-step}
\end{figure}
